---
title: "All MS Risk Variants"
format: html
editor: visual
date: 2026-02-25
author:
  - name: Ellie Kim
    email: ellie.kim@cchmc.org
    affiliation:
      - name: CCHMC
---

```{r}

library(readxl)
library(writexl)
library(dplyr)
library(openxlsx)
library(stringr)
library(chromoMap)
library(rtracklayer)
library(stringr)
library(readr)
library(tidyr)

```

## Getting RELI Pre-Processing

So far, we have a list of tag risk variants from curated GWAS studies. We need to perform LD-pruning to get all independent risk loci.

```{r}
RELI_prep <- read_excel("gwas-association-downloaded_2026-02-13-MONDO_0005301-withChildTraits.xlsx", sheet = "RELI_input")

write.table(RELI_prep, 
            file = "reli_input_mock.txt", 
            sep = "\t",             # Use tabs (common for .txt files)
            row.names = FALSE,      # Prevents R from adding a column of row numbers
            quote = FALSE)          # Removes quotes around text values (cleaner look)
```

Run RELI LD pruning

``` unix
{r}
module load reli/0.1.1a

reli-ld-prune \
  -in reli_input.txt \
  -ancestry EUR \
  -build hg38
```

Output of RELI LD pruning: in.snp.LD-pvalue-pruned and snp.missing (not found in PLINK database)

Question: Add those missing snps to the chromomap or not? Will make 2 versions. Note that this includes all the risk alleles from the GWAS catalog (including those tag variants I could not find the risk allele for).

## Chromomap

Chromomap of LD-pruned SNPs

```{r}
LD_pruned <- read.delim("in.snp.LD-pvalue-pruned", header = TRUE, sep = '\t')
reference <- read_excel("gwas-association-downloaded_2026-02-13-MONDO_0005301-withChildTraits.xlsx", sheet = "Included_SNPs")
reference_unique <- reference %>% distinct(SNPS, .keep_all = TRUE) # this is to prevent duplicate rsids. 

library(dplyr)
chromomap_table <- LD_pruned %>%
  # Convert chr to "chrN"
  mutate(chrN = paste0("chr", chr)) %>%
  
  # Join with reference table by rsid
  left_join(reference_unique, by = c("rsID" = "SNPS")) %>%
  
  # Create start and end positions
  mutate(
    start_pos = CHR_POS,
    end_pos = CHR_POS + 1,
    label = "GWAS_Tag_Variant"
  ) %>%
  
  # Select only required columns
  select(rsID, chrN, start_pos, end_pos, label)

write.table(chromomap_table, file = "chromomap_input_LDpruned.txt", sep = "\t", row.names = FALSE, col.names = FALSE, quote = FALSE )

chromoMap(
  "hg38_chromosomes.txt",
  "chromomap_input_LDpruned.txt",
  chr_color = c("black"),
  data_based_color_map = FALSE,
  data_type = "categorical",
  data_colors = list(c("#6BAED6")),
  legend = TRUE,
  canvas_width = 900,
  canvas_height = 920,
  export.options = TRUE
)
```

Chromomap of LD-pruned SNPs + missing SNPs

```{r}
snp_missing <- read.delim("snp.missing", header = FALSE, sep = "\t")
chromomap_table_missing <- snp_missing %>%
  # Join with reference table by rsid
  left_join(reference_unique, by = c("V1" = "SNPS")) %>%
  
  # Create start and end positions
  mutate(
    chrN = paste("chr", CHR_ID),
    start_pos = CHR_POS,
    end_pos = CHR_POS + 1,
    label = "GWAS_Tag_Variant"
  ) %>%
  
  # Select only required columns
  select(V1, chrN, start_pos, end_pos, label)

write.table(chromomap_table_missing, file = "chromomap_input_snp_missing.txt", sep = "\t", row.names = FALSE, col.names = FALSE, quote = FALSE )

# join both tables - LD pruned + snp missing
pruned <- read.delim("chromomap_input_LDpruned.txt", header = FALSE, sep = '\t')
missing <- read.delim("chromomap_input_snp_missing.txt", header = FALSE, sep = '\t')
combined <- bind_rows(pruned, missing)
write.table(combined, file = "chromomap_input_combinedLD+missing.txt", sep = "\t", row.names = FALSE, col.names = FALSE, quote = FALSE )


chromoMap(
  "hg38_chromosomes.txt",
  "chromomap_input_combinedLD+missing.txt",
  chr_color = c("black"),
  data_based_color_map = FALSE,
  data_type = "categorical",
  data_colors = list(c("#6BAED6")),
  legend = TRUE,
  canvas_width = 900,
  canvas_height = 920,
  export.options = TRUE
)
```

LD Expansion

``` unix
{r}
module load ez_pipeline/0.8

ez_pipeline -input in.snp.LD-pvalue-pruned -dir . -ldexpansion -ancestry EUR -build hg38

bash step_1.bat
```

Now, we have the tag snp and the expanded snp, along with the R2. Let's build a table with all the information. Going forward, tables should only have **6152 rows**. (now 6150 after de-duplication).

```{r}
LD_pairs <- read.table("./LDexpansion/snp_expanded.bed", header = FALSE, fill = TRUE, stringsAsFactors = FALSE)
LD_pairs <- LD_pairs[LD_pairs$V1 != "CHR_A", ]
colnames(LD_pairs) <- c("CHR_TAG", "POS_TAG", "SNP_TAG", "CHR_EXP", "BP_EXP", "SNP_EXP", "R2")
LD_pairs <- LD_pairs %>%
  select("SNP_TAG", "CHR_TAG", "POS_TAG", "SNP_EXP", "CHR_EXP", "BP_EXP", "R2")

write.xlsx(LD_pairs, file = "MS_Risk_SNPs.xlsx", rowNames = FALSE)

```

## Converting chr pos to rsID

RELI gave us a mix of rsID and chromosome positions after LD expansion. We will convert chromosome positions to rsID using UCSC snp tables. First, we convert the big bed file to a bed file.

``` unix
{r}
bigBedToBed dbSnp155Common.bb dbSnp155Common.bed
```

```{r}
dbsnp_df <- read.table( "dbSnp155Common_minimal.bed", sep = "\t", header = FALSE, stringsAsFactors = FALSE ) 

colnames(dbsnp_df) <- c("chrom", "chromStart", "chromEnd", "rsid", "ref", "alts")

# extract chr:pos entries from the SNP_EXP column. 
df_chrpos <- LD_pairs %>% 
  filter(grepl("^chr[0-9XYM]+:[0-9]+", SNP_EXP)) %>% 
  mutate( 
    chrom = str_extract(SNP_EXP, "^chr[0-9XYM]+"), 
    pos = as.integer(str_extract(SNP_EXP, "(?<=:)[0-9]+")) 
    )

# join with the dbSNP df by chrom and chromStart
df_chrpos_joined <- df_chrpos %>% 
  left_join( 
    dbsnp_df, by = c("chrom" = "chrom", "pos" = "chromStart") 
    )

# this created more rows, so that means there are duplicated chr positions that matched to multiple rsIDs. Let's handle these multi-hit positions by collapsing rsIDs. 
df_chrpos_joined <- df_chrpos_joined %>% 
  group_by(SNP_EXP) %>% 
  summarise( rsid = paste(unique(rsid), collapse = ","), .groups = "drop" )

# replace chr:pos entries in the original df. 
LD_pairs_fixed <- LD_pairs %>% 
  left_join(df_chrpos_joined, by = "SNP_EXP") %>% 
  mutate(SNP_EXP = ifelse(!is.na(rsid), rsid, SNP_EXP)) %>% 
  select(-rsid)

wb <- loadWorkbook("MS_Risk_SNPs.xlsx")
addWorksheet(wb, "SNP_chrtorsid")
writeData(wb, "SNP_chrtorsid", LD_pairs_fixed)

# Save back to the same file
saveWorkbook(
  wb,
  "MS_Risk_SNPs.xlsx",
  overwrite = TRUE
)
```

## Add Ancestry annotation

Ancestry annotations need to be added prior to frequency since we want ancestry specific frequency information.

```{r}
ref_ancestry <- reference %>% 
  select(SNPS, `INITIAL SAMPLE SIZE`) %>% 
  distinct(SNPS, .keep_all = TRUE)

LD_pairs_ancestry <- LD_pairs_fixed %>% 
  left_join(ref_ancestry, by = c("SNP_TAG" = "SNPS")) %>% 
  mutate(
    Ancestry = case_when(
      str_detect(`INITIAL SAMPLE SIZE`, regex("Taiwan", ignore_case = TRUE)) ~ "EAS",
      str_detect(`INITIAL SAMPLE SIZE`, regex("African", ignore_case = TRUE)) ~ "AFR",
      TRUE ~ "EUR"
    )
  ) %>% 
  select(-`INITIAL SAMPLE SIZE`)

wb <- loadWorkbook("MS_Risk_SNPs.xlsx")
addWorksheet(wb, "SNP_ancestry")
writeData(wb, "SNP_ancestry", LD_pairs_ancestry)

# Save back to the same file
saveWorkbook(
  wb,
  "MS_Risk_SNPs.xlsx",
  overwrite = TRUE
)
```

## Ref/Alt Alleles

Let's also add the reference and alt alleles. This seems a little round-about because making the LD_pairs_with_refalt made a lot of duplicated rows, so we used the results from that ref/alt to make our reference ref/alt columns.

```{r}
dbsnp_refalt <- dbsnp_df %>%
  select(rsid, ref, alts)

LD_pairs_with_refalt <- LD_pairs_ancestry %>% left_join(dbsnp_refalt, by = c("SNP_EXP" = "rsid"))

collapsed_refalt <- LD_pairs_with_refalt %>%
  group_by(SNP_EXP) %>%
  summarise(
    ref  = paste(unique(ref), collapse = ","),
    alts = paste(unique(alts), collapse = ","),
    .groups = "drop"
  )

LD_pairs_final_refalt <- LD_pairs_ancestry %>% left_join(collapsed_refalt, by = "SNP_EXP")

wb <- loadWorkbook("MS_Risk_SNPs.xlsx")
addWorksheet(wb, "SNP_refalt")
writeData(wb, "SNP_refalt", LD_pairs_final_refalt)

# Save back to the same file
saveWorkbook(
  wb,
  "MS_Risk_SNPs.xlsx",
  overwrite = TRUE
)
```

## PLINK —freq

To obtain ancestry-specific frequency information, we use PLINK.

``` unix
{r}
awk '$3=="AFR"{print $1}' integrated_call_samples_v3.20130502.ALL.panel > afr_samples.txt
awk '$3=="EUR"{print $1}' integrated_call_samples_v3.20130502.ALL.panel > eur_samples.txt

# edit this file manually to duplicate the column. I named this file AFR_keep.txt and EUR_keep.txt
```

Split the snps of interest by Ancestry.

```{r}
SNPs <- read_excel("MS_Risk_SNPs.xlsx", sheet = "SNP_refalt")

EUR_snps <- SNPs %>% 
  filter(Ancestry == "EUR", !is.na(SNP_EXP)) %>% 
  pull(SNP_EXP) %>% 
  unique()

write.table(EUR_snps, file = "EUR_snps_of_interest.txt", quote = FALSE, row.names = FALSE, col.names = FALSE)

AFR_snps <- SNPs %>% 
  filter(Ancestry == "AFR",!is.na(SNP_EXP)) %>% 
  pull(SNP_EXP) %>% 
  unique()

write.table(AFR_snps, file = "AFR_snps_of_interest.txt", quote = FALSE, row.names = FALSE, col.names = FALSE)

# these files will be moved to the PLINK folder for plink --freq functions. 
```

Run PLINK (used /1.90b)

``` unix
{r}
for chr in {1..22}; do
  plink \
    --bfile chr${chr}_1000G \
    --keep EUR_keep.txt \
    --extract EUR_snps_of_interest.txt \
    --freq \
    --out eur_freq_chr${chr}
done

for chr in {1..22}; do
  plink \
    --bfile chr${chr}_1000G \
    --keep AFR_keep.txt \
    --extract AFR_snps_of_interest.txt \
    --freq \
    --out afr_freq_chr${chr}
done
```

Integrating PLINK results: First, we get the .frq files into a readable format.

```{r}
eur_freq <- lapply(1:22, function(chr) {
  read_table(
    paste0("../plink/eur_freq_chr", chr, ".frq"),
    col_types = cols(
      CHR      = col_integer(),
      SNP      = col_character(),
      A1       = col_character(),
      A2       = col_character(),
      MAF      = col_double(),
      NCHROBS  = col_double()
    )
  )
}) %>%
  bind_rows()

eur_freq <- eur_freq %>%
  mutate(
    minor_allele = A1,
    major_allele = A2,
    minor_af = MAF,
    major_af = 1 - MAF
  )
eur_freq <- subset(eur_freq, select = -c(A1, A2, MAF, NCHROBS))
eur_freq_unique <- eur_freq %>%
  group_by(SNP) %>%
  slice(1) %>%  # keep the first occurrence
  ungroup()

afr_freq <- lapply(c(5,20), function(chr) {
  read_table(
    paste0("../plink/afr_freq_chr", chr, ".frq"),
    col_types = cols(
      CHR      = col_integer(),
      SNP      = col_character(),
      A1       = col_character(),
      A2       = col_character(),
      MAF      = col_double(),
      NCHROBS  = col_double()
    )
  )
}) %>%
  bind_rows()

afr_freq <- afr_freq %>%
  mutate(
    minor_allele = A1,
    major_allele = A2,
    minor_af = MAF,
    major_af = 1 - MAF
  )
afr_freq <- subset(afr_freq, select = -c(A1, A2, MAF, NCHROBS))
```

Adding to excel sheet

```{r}
SNPs_frq <- SNPs %>%
  mutate(
    snp_maj = NA_character_,
    snp_min = NA_character_,
    snp_maj_af = NA_real_,
    snp_min_af = NA_real_,
    tag_maj = NA_character_,
    tag_min = NA_character_,
    tag_maj_af = NA_real_,
    tag_min_af = NA_real_
  )

SNPs_frq <- SNPs_frq %>% 
  left_join( 
    eur_freq %>% 
      select(SNP, major_allele, minor_allele, major_af, minor_af), 
    by = c("SNP_EXP" = "SNP") ) %>% 
  mutate( 
    snp_maj = ifelse(Ancestry == "EUR", major_allele, snp_maj), 
    snp_min = ifelse(Ancestry == "EUR", minor_allele, snp_min), 
    snp_maj_af = ifelse(Ancestry == "EUR", major_af, snp_maj_af), 
    snp_min_af = ifelse(Ancestry == "EUR", minor_af, snp_min_af) ) %>% 
  select(-major_allele, -minor_allele, -major_af, -minor_af)

SNPs_frq <- SNPs_frq %>% 
  left_join( 
    afr_freq %>% 
      select(SNP, major_allele, minor_allele, major_af, minor_af), 
    by = c("SNP_EXP" = "SNP") ) %>% 
  mutate( 
    snp_maj = ifelse(Ancestry == "AFR", major_allele, snp_maj), 
    snp_min = ifelse(Ancestry == "AFR", minor_allele, snp_min), 
    snp_maj_af = ifelse(Ancestry == "AFR", major_af, snp_maj_af), 
    snp_min_af = ifelse(Ancestry == "AFR", minor_af, snp_min_af) ) %>% 
  select(-major_allele, -minor_allele, -major_af, -minor_af)


# Adding in tag variant info
SNPs_frq <- SNPs_frq %>% 
  left_join( 
    eur_freq %>% 
      select(SNP, major_allele, minor_allele, major_af, minor_af), 
    by = c("SNP_TAG" = "SNP") ) %>% 
  mutate( 
    tag_maj = ifelse(Ancestry == "EUR", major_allele, tag_maj), 
    tag_min = ifelse(Ancestry == "EUR", minor_allele, tag_min), 
    tag_maj_af = ifelse(Ancestry == "EUR", major_af, tag_maj_af), 
    tag_min_af = ifelse(Ancestry == "EUR", minor_af, tag_min_af) ) %>% 
  select(-major_allele, -minor_allele, -major_af, -minor_af)

SNPs_frq <- SNPs_frq %>% 
  left_join( 
    afr_freq %>% 
      select(SNP, major_allele, minor_allele, major_af, minor_af), 
    by = c("SNP_TAG" = "SNP") ) %>% 
  mutate( 
    tag_maj = ifelse(Ancestry == "AFR", major_allele, tag_maj), 
    tag_min = ifelse(Ancestry == "AFR", minor_allele, tag_min), 
    tag_maj_af = ifelse(Ancestry == "AFR", major_af, tag_maj_af), 
    tag_min_af = ifelse(Ancestry == "AFR", minor_af, tag_min_af) ) %>% 
  select(-major_allele, -minor_allele, -major_af, -minor_af)



# make sure to take care of duplicate snps (removed manually in excel)
dup_snps <- SNPs_frq %>% count(SNP_EXP) %>% filter(n > 1)

wb <- loadWorkbook("MS_Risk_SNPs.xlsx")
addWorksheet(wb, "SNPS_frq")
writeData(wb, "SNPS_frq", SNPs_frq)

# Save back to the same file
saveWorkbook(
  wb,
  "MS_Risk_SNPs.xlsx",
  overwrite = TRUE
)
```

## Risk Alleles

```{r}
RiskAlleles <- read_excel("gwas-association-downloaded_2026-02-13-MONDO_0005301-withChildTraits.xlsx", sheet = "Tag_Risk_Alleles")

SNPs <- read_excel("MS_Risk_SNPs.xlsx", sheet = "SNPs_frq")

SNPs_risk <- SNPs %>%
  
  # 1. Join risk alleles by SNP_TAG
  left_join(
    RiskAlleles %>% select(SNP, Risk),
    by = c("SNP_TAG" = "SNP")
  ) %>%
  
  # 2. Create tag_risk (Risk column from RiskAlleles)
  mutate(
    tag_risk = Risk
  ) %>%
  
  # 3. Assign snp_risk with full error handling
  mutate(
    snp_risk = case_when(
      # No reported risk allele for this SNP_TAG
      is.na(tag_risk) ~ "ERROR: no reported tag risk allele",
      
      # tag_risk matches tag major allele
      tag_risk == tag_maj ~ snp_maj,
      
      # tag_risk matches tag minor allele
      tag_risk == tag_min ~ snp_min,
      
      # tag_risk exists but does not match either allele
      TRUE ~ "ERROR: risk allele does not match"
    )
  ) %>%
  
  # 4. Remove the Risk column (optional)
  select(-Risk)

wb <- loadWorkbook("MS_Risk_SNPs.xlsx")
addWorksheet(wb, "SNPs_risk")
writeData(wb, "SNPs_risk", SNPs_risk)

# Save back to the same file
saveWorkbook(
  wb,
  "MS_Risk_SNPs.xlsx",
  overwrite = TRUE
)
```

So far, 793 SNPs are missing frequency information (not in 1000G vX database) so they do not have risk allele information. This is 793/6150 SNPs = 13% of the SNPs.
