---
title: "All MS Risk Variants"
format: html
editor: visual
date: 2026-02-25
author:
  - name: Ellie Kim
    email: ellie.kim@cchmc.org
    affiliation:
      - name: CCHMC
---

**Methods**

*Curation of MS Risk SNPs* 

Summary statistics for the trait “multiple sclerosis” were downloaded from the GWAS Catalog (doi.org/10.1093/nar/gkae1070) on February, 2026. Because results included MS-like conditions like neuromyelitis optica, results were further filtered to include only the multiple sclerosis trait. SNPs meeting GWAS significance at p \< 5e-8 were included. Our inclusion criteria for curating MS risk SNPs were studies that compared a group of MS patients to a cohort of healthy controls, and these studies are listed in Table S1. Therefore, studies that compared traits within the MS cohort (i.e., antibody status, drug response), and studies that compared MS to another trait (i.e. MS compared to cardiovascular disease or cancers) were excluded.  

*Annotation of LD-Expanded Risk SNPs*  

To nominate unique risk loci, the list of GWAS-nominated tag SNPs were LD-pruned using PLINK (doi.org/10.1086/519795) with a cutoff of R\^2 \> 0.8. These risk loci were then LD-expanded using PLINK and the 1000Genomes phase 3 database (doi.org/10.1038/nature15393). The expanded SNPs that were not given a rsID were labeled with one, if applicable, using UCSC SNP table dbSnp155 (doi.org/10.1093/nar/gky1048). Allele frequencies for tag and expanded SNPs were computed using PLINK and the 1000Genome phase 3 database. SNPs that were not found in this database were supplemented with information from the 1000Genomes 30X coverage database (doi.org/10.1016/j.cell.2022.08.004). Risk allele annotations were assigned based on whether the given risk allele corresponded to the major or minor allele. For example, if the risk allele of a tag variant is A which happens to be the major allele, the risk allele of the expanded variant is the major allele. In cases where the given risk allele did not correspond to either major or minor allele, the risk allele was manually cross checked with the original manuscript. Gene annotations for each SNP were obtained from SNPNexus (doi.org/10.1093/nar/gkaa420). 

```{r}
library(readxl)
library(writexl)
library(dplyr)
library(openxlsx)
library(stringr)
library(chromoMap)
library(rtracklayer)
library(stringr)
library(readr)
library(tidyr)
```

## Getting RELI Pre-Processing

So far, we have a list of tag risk variants from curated GWAS studies. We need to perform LD-pruning to get all independent risk loci.

```{r}
RELI_prep <- read_excel("gwas-association-downloaded_2026-02-13-MONDO_0005301-withChildTraits.xlsx", sheet = "RELI_input")

write.table(RELI_prep, 
            file = "reli_input.txt",
            sep = "\t",
            row.names = FALSE,
            quote = FALSE)  
```

Run [RELI](https://github.com/weirauchlab/reli#how-to-cite) LD pruning (https://doi.org/10.1038/s41588-018-0102-3)

RELI has pre-processing steps (LD pruning and LD expansion) that takes a list of SNPs and "prunes" it down to unique loci. Then, we perform LD-expansion to get the list of all SNPs that are in the same LD-block as the unique loci, giving us the broadest coverage of all potential MS risk variants.

``` unix

module load reli/0.1.1a

reli-ld-prune \
  -in reli_input.txt \
  -ancestry EUR \
  -build hg38
```

Relevant outputs of RELI LD pruning: in.snp.LD-pvalue-pruned and snp.missing (not found in PLINK 1000G phase 3 database)

## Chromomap

This section creates a chromomap, which is a visual representation of the independent risk loci across the human genome.

Chromomap of LD-pruned SNPs

```{r}
LD_pruned <- read.delim("in.snp.LD-pvalue-pruned", header = TRUE, sep = '\t')
reference <- read_excel("gwas-association-downloaded_2026-02-13-MONDO_0005301-withChildTraits.xlsx", sheet = "Included_SNPs")
reference_unique <- reference %>% distinct(SNPS, .keep_all = TRUE) # this is to prevent duplicate rsids. 

chromomap_table <- LD_pruned %>%
  # Convert chr to "chrN"
  mutate(chrN = paste0("chr", chr)) %>%
  
  left_join(reference_unique, by = c("rsID" = "SNPS")) %>%
  
  # Create start and end positions
  mutate(
    start_pos = CHR_POS,
    end_pos = CHR_POS + 1,
    label = "GWAS_Tag_Variant"
  ) %>%
  
  # Select only required columns
  select(rsID, chrN, start_pos, end_pos, label)

write.table(chromomap_table, file = "chromomap_input_LDpruned.txt", sep = "\t", row.names = FALSE, col.names = FALSE, quote = FALSE )

chromoMap(
  "hg38_chromosomes.txt",
  "chromomap_input_LDpruned.txt",
  chr_color = c("black"),
  data_based_color_map = FALSE,
  data_type = "categorical",
  data_colors = list(c("#6BAED6")),
  legend = TRUE,
  canvas_width = 900,
  canvas_height = 920,
  export.options = TRUE
)
```

Chromomap of LD-pruned SNPs + missing SNPs

```{r}
snp_missing <- read.delim("snp.missing", header = FALSE, sep = "\t")
chromomap_table_missing <- snp_missing %>%
  # Join with reference table by rsid
  left_join(reference_unique, by = c("V1" = "SNPS")) %>%
  
  # Create start and end positions
  mutate(
    chrN = paste("chr", CHR_ID),
    start_pos = CHR_POS,
    end_pos = CHR_POS + 1,
    label = "GWAS_Tag_Variant"
  ) %>%
  
  # Select only required columns
  select(V1, chrN, start_pos, end_pos, label)

write.table(chromomap_table_missing, file = "chromomap_input_snp_missing.txt", sep = "\t", row.names = FALSE, col.names = FALSE, quote = FALSE )

# join both tables - LD pruned + snp missing
pruned <- read.delim("chromomap_input_LDpruned.txt", header = FALSE, sep = '\t')
missing <- read.delim("chromomap_input_snp_missing.txt", header = FALSE, sep = '\t')
combined <- bind_rows(pruned, missing)
write.table(combined, file = "chromomap_input_combinedLD+missing.txt", sep = "\t", row.names = FALSE, col.names = FALSE, quote = FALSE )


chromoMap(
  "hg38_chromosomes.txt",
  "chromomap_input_combinedLD+missing.txt",
  chr_color = c("black"),
  data_based_color_map = FALSE,
  data_type = "categorical",
  data_colors = list(c("#6BAED6")),
  legend = TRUE,
  canvas_width = 900,
  canvas_height = 920,
  export.options = TRUE
)
```

## LD Expansion from pruned list of rsids

``` unix

module load ez_pipeline/0.8

ez_pipeline -input in.snp.LD-pvalue-pruned -dir . -ldexpansion -ancestry EUR -build hg38

bash step_1.bat
```

Now, we have the tag snp and the expanded snp, along with the R2. Let's build a table with all the information.

```{r}
LD_pairs <- read.table("./LDexpansion/snp_expanded.bed", header = FALSE, fill = TRUE, stringsAsFactors = FALSE)
LD_pairs <- LD_pairs[LD_pairs$V1 != "CHR_A", ]
colnames(LD_pairs) <- c("CHR_TAG", "POS_TAG", "SNP_TAG", "CHR_EXP", "BP_EXP", "SNP_EXP", "R2")
LD_pairs <- LD_pairs %>%
  select("SNP_TAG", "CHR_TAG", "POS_TAG", "SNP_EXP", "CHR_EXP", "BP_EXP", "R2")

write.xlsx(LD_pairs, file = "MS_Risk_SNPs.xlsx", rowNames = FALSE)

```

## Converting chr pos to rsID

RELI gave us a mix of rsID and chromosome positions after LD expansion. We will convert chromosome positions to rsID using UCSC snp tables. First, we convert the big bed file to a bed file. The bigger bed file was minimized to a "minimal" bed with just the information we need for downstream processing.

``` unix

bigBedToBed dbSnp155Common.bb dbSnp155Common.bed
cut -f1-5,7 dbSnp155Common.bed > dbSnp155Common_minimal.bed
```

```{r}
dbsnp_df <- read.table( "dbSnp155Common_minimal.bed", sep = "\t", header = FALSE, stringsAsFactors = FALSE ) 

colnames(dbsnp_df) <- c("chrom", "chromStart", "chromEnd", "rsid", "ref", "alts")

# extract chr:pos entries from the SNP_EXP column. 
df_chrpos <- LD_pairs %>% 
  filter(grepl("^chr[0-9XYM]+:[0-9]+", SNP_EXP)) %>% 
  mutate( 
    chrom = str_extract(SNP_EXP, "^chr[0-9XYM]+"), 
    pos = as.integer(str_extract(SNP_EXP, "(?<=:)[0-9]+")) 
    )

# join with the dbSNP df by chrom and chromStart
df_chrpos_joined <- df_chrpos %>% 
  left_join( 
    dbsnp_df, by = c("chrom" = "chrom", "pos" = "chromStart") 
    )

# this created more rows, so that means there are duplicated chr positions that matched to multiple rsIDs. Let's handle these multi-hit positions by collapsing rsIDs. 
df_chrpos_joined <- df_chrpos_joined %>% 
  group_by(SNP_EXP) %>% 
  summarise( rsid = paste(unique(rsid), collapse = ","), .groups = "drop" )

# replace chr:pos entries in the original df. 
LD_pairs_fixed <- LD_pairs %>% 
  left_join(df_chrpos_joined, by = "SNP_EXP") %>% 
  mutate(SNP_EXP = ifelse(!is.na(rsid), rsid, SNP_EXP)) %>% 
  select(-rsid)

wb <- loadWorkbook("MS_Risk_SNPs.xlsx")
addWorksheet(wb, "SNP_chrtorsid")
writeData(wb, "SNP_chrtorsid", LD_pairs_fixed)

# Save back to the same file
saveWorkbook(
  wb,
  "MS_Risk_SNPs.xlsx",
  overwrite = TRUE
)
```

## Add Ancestry annotation

Ancestry annotations need to be added prior to frequency since we want ancestry specific frequency information.

```{r}
ref_ancestry <- reference %>% 
  select(SNPS, `INITIAL SAMPLE SIZE`) %>% 
  distinct(SNPS, .keep_all = TRUE)

LD_pairs_ancestry <- LD_pairs_fixed %>% 
  left_join(ref_ancestry, by = c("SNP_TAG" = "SNPS")) %>% 
  mutate(
    Ancestry = case_when(
      str_detect(`INITIAL SAMPLE SIZE`, regex("Taiwan", ignore_case = TRUE)) ~ "EAS",
      str_detect(`INITIAL SAMPLE SIZE`, regex("African", ignore_case = TRUE)) ~ "AFR",
      TRUE ~ "EUR"
    )
  ) %>% 
  select(-`INITIAL SAMPLE SIZE`)

wb <- loadWorkbook("MS_Risk_SNPs.xlsx")
addWorksheet(wb, "SNP_ancestry")
writeData(wb, "SNP_ancestry", LD_pairs_ancestry)

# Save back to the same file
saveWorkbook(
  wb,
  "MS_Risk_SNPs.xlsx",
  overwrite = TRUE
)
```

## Ref/Alt Alleles

Let's also add the reference and alt alleles. This seems a little round-about because making the LD_pairs_with_refalt made a lot of duplicated rows, so we used the results from that ref/alt to make our reference ref/alt columns.

```{r}
dbsnp_refalt <- dbsnp_df %>%
  select(rsid, ref, alts)

LD_pairs_with_refalt <- LD_pairs_ancestry %>% left_join(dbsnp_refalt, by = c("SNP_EXP" = "rsid"))

collapsed_refalt <- LD_pairs_with_refalt %>%
  group_by(SNP_EXP) %>%
  summarise(
    ref  = paste(unique(ref), collapse = ","),
    alts = paste(unique(alts), collapse = ","),
    .groups = "drop"
  )

LD_pairs_final_refalt <- LD_pairs_ancestry %>% left_join(collapsed_refalt, by = "SNP_EXP")

wb <- loadWorkbook("MS_Risk_SNPs.xlsx")
addWorksheet(wb, "SNP_refalt")
writeData(wb, "SNP_refalt", LD_pairs_final_refalt)

# Save back to the same file
saveWorkbook(
  wb,
  "MS_Risk_SNPs.xlsx",
  overwrite = TRUE
)
```

## PLINK —freq

To obtain ancestry-specific frequency information, we use PLINK, which uses the 1000Genomes phase 3 database.

``` unix

awk '$3=="AFR"{print $1 "\t" $1}' integrated_call_samples_v3.20130502.ALL.panel > AFR_keep.txt
awk '$3=="EUR"{print $1 "\t" $1}' integrated_call_samples_v3.20130502.ALL.panel > EUR_keep.txt
```

Split the SNPs of interest by Ancestry.

```{r}
SNPs <- read_excel("MS_Risk_SNPs.xlsx", sheet = "SNP_refalt")

EUR_snps <- SNPs %>% 
  filter(Ancestry == "EUR", !is.na(SNP_EXP)) %>% 
  pull(SNP_EXP) %>% 
  unique()

write.table(EUR_snps, file = "EUR_snps_of_interest.txt", quote = FALSE, row.names = FALSE, col.names = FALSE)

AFR_snps <- SNPs %>% 
  filter(Ancestry == "AFR",!is.na(SNP_EXP)) %>% 
  pull(SNP_EXP) %>% 
  unique()

write.table(AFR_snps, file = "AFR_snps_of_interest.txt", quote = FALSE, row.names = FALSE, col.names = FALSE)

# these files will be moved to the PLINK folder for plink --freq functions. 
```

Run PLINK (v.1.90b)

``` unix

for chr in {1..22}; do
  plink \
    --bfile chr${chr}_1000G \
    --keep EUR_keep.txt \
    --extract EUR_snps_of_interest.txt \
    --freq \
    --out eur_freq_chr${chr}
done

for chr in {1..22}; do
  plink \
    --bfile chr${chr}_1000G \
    --keep AFR_keep.txt \
    --extract AFR_snps_of_interest.txt \
    --freq \
    --out afr_freq_chr${chr}
done
```

Integrating PLINK results: First, we get the .frq files into a readable format.

```{r}
eur_freq <- lapply(1:22, function(chr) {
  read_table(
    paste0("../plink/eur_freq_chr", chr, ".frq"),
    col_types = cols(
      CHR      = col_integer(),
      SNP      = col_character(),
      A1       = col_character(),
      A2       = col_character(),
      MAF      = col_double(),
      NCHROBS  = col_double()
    )
  )
}) %>%
  bind_rows()

eur_freq <- eur_freq %>%
  mutate(
    minor_allele = A1,
    major_allele = A2,
    minor_af = MAF,
    major_af = 1 - MAF
  )
eur_freq <- subset(eur_freq, select = -c(A1, A2, MAF, NCHROBS))
eur_freq_unique <- eur_freq %>%
  group_by(SNP) %>%
  slice(1) %>%  # keep the first occurrence
  ungroup()

afr_freq <- lapply(c(5,20), function(chr) {
  read_table(
    paste0("../plink/afr_freq_chr", chr, ".frq"),
    col_types = cols(
      CHR      = col_integer(),
      SNP      = col_character(),
      A1       = col_character(),
      A2       = col_character(),
      MAF      = col_double(),
      NCHROBS  = col_double()
    )
  )
}) %>%
  bind_rows()

afr_freq <- afr_freq %>%
  mutate(
    minor_allele = A1,
    major_allele = A2,
    minor_af = MAF,
    major_af = 1 - MAF
  )
afr_freq <- subset(afr_freq, select = -c(A1, A2, MAF, NCHROBS))
```

Adding to excel sheet

```{r}
SNPs_frq <- SNPs %>%
  mutate(
    snp_maj = NA_character_,
    snp_min = NA_character_,
    snp_maj_af = NA_real_,
    snp_min_af = NA_real_,
    tag_maj = NA_character_,
    tag_min = NA_character_,
    tag_maj_af = NA_real_,
    tag_min_af = NA_real_
  )

SNPs_frq <- SNPs_frq %>% 
  left_join( 
    eur_freq %>% 
      select(SNP, major_allele, minor_allele, major_af, minor_af), 
    by = c("SNP_EXP" = "SNP") ) %>% 
  mutate( 
    snp_maj = ifelse(Ancestry == "EUR", major_allele, snp_maj), 
    snp_min = ifelse(Ancestry == "EUR", minor_allele, snp_min), 
    snp_maj_af = ifelse(Ancestry == "EUR", major_af, snp_maj_af), 
    snp_min_af = ifelse(Ancestry == "EUR", minor_af, snp_min_af) ) %>% 
  select(-major_allele, -minor_allele, -major_af, -minor_af)

SNPs_frq <- SNPs_frq %>% 
  left_join( 
    afr_freq %>% 
      select(SNP, major_allele, minor_allele, major_af, minor_af), 
    by = c("SNP_EXP" = "SNP") ) %>% 
  mutate( 
    snp_maj = ifelse(Ancestry == "AFR", major_allele, snp_maj), 
    snp_min = ifelse(Ancestry == "AFR", minor_allele, snp_min), 
    snp_maj_af = ifelse(Ancestry == "AFR", major_af, snp_maj_af), 
    snp_min_af = ifelse(Ancestry == "AFR", minor_af, snp_min_af) ) %>% 
  select(-major_allele, -minor_allele, -major_af, -minor_af)


# Adding in tag variant info
SNPs_frq <- SNPs_frq %>% 
  left_join( 
    eur_freq %>% 
      select(SNP, major_allele, minor_allele, major_af, minor_af), 
    by = c("SNP_TAG" = "SNP") ) %>% 
  mutate( 
    tag_maj = ifelse(Ancestry == "EUR", major_allele, tag_maj), 
    tag_min = ifelse(Ancestry == "EUR", minor_allele, tag_min), 
    tag_maj_af = ifelse(Ancestry == "EUR", major_af, tag_maj_af), 
    tag_min_af = ifelse(Ancestry == "EUR", minor_af, tag_min_af) ) %>% 
  select(-major_allele, -minor_allele, -major_af, -minor_af)

SNPs_frq <- SNPs_frq %>% 
  left_join( 
    afr_freq %>% 
      select(SNP, major_allele, minor_allele, major_af, minor_af), 
    by = c("SNP_TAG" = "SNP") ) %>% 
  mutate( 
    tag_maj = ifelse(Ancestry == "AFR", major_allele, tag_maj), 
    tag_min = ifelse(Ancestry == "AFR", minor_allele, tag_min), 
    tag_maj_af = ifelse(Ancestry == "AFR", major_af, tag_maj_af), 
    tag_min_af = ifelse(Ancestry == "AFR", minor_af, tag_min_af) ) %>% 
  select(-major_allele, -minor_allele, -major_af, -minor_af)



# Take care of duplicate SNPs in dup_snps manually in Excel (we kept the duplicate with frequency that was closest to the corresponding tag SNP)
dup_snps <- SNPs_frq %>% count(SNP_EXP) %>% filter(n > 1)

wb <- loadWorkbook("MS_Risk_SNPs.xlsx")
addWorksheet(wb, "SNPS_frq")
writeData(wb, "SNPS_frq", SNPs_frq)

# Save back to the same file
saveWorkbook(
  wb,
  "MS_Risk_SNPs.xlsx",
  overwrite = TRUE
)
```

## Risk Alleles

In this section, we are now annotating each SNP with the MS risk allele. The tag variant risk allele has been obtained from the GWAS catalog and by cross-referencing the manuscript the variant came from. These are logged in the sheet "Tag_Risk_Alleles" which will be used as the reference. Note that this reference was manually curated from the GWAS catalog and the supplemental files of the studies in the GWAS catalog.

```{r}
RiskAlleles <- read_excel("gwas-association-downloaded_2026-02-13-MONDO_0005301-withChildTraits.xlsx", sheet = "Tag_Risk_Alleles")

SNPs <- read_excel("MS_Risk_SNPs.xlsx", sheet = "SNPs_frq")

SNPs_risk <- SNPs %>%
    left_join(
    RiskAlleles %>% select(SNP, Risk),
    by = c("SNP_TAG" = "SNP")
  ) %>%
    mutate(
    tag_risk = Risk
  ) %>%
  
  # assign snp_risk with error messages
  mutate(
    snp_risk = case_when(
      # No reported risk allele for this SNP_TAG
      is.na(tag_risk) ~ "ERROR: no reported tag risk allele",
      # tag_risk matches tag major allele
      tag_risk == tag_maj ~ snp_maj,
      # tag_risk matches tag minor allele
      tag_risk == tag_min ~ snp_min,
      # tag_risk exists but does not match either allele
      TRUE ~ "ERROR: risk allele does not match"
    )
  ) %>%
  
  # remove the Risk column since it was only used as a reference
  select(-Risk)

wb <- loadWorkbook("MS_Risk_SNPs.xlsx")
addWorksheet(wb, "SNPs_risk")
writeData(wb, "SNPs_risk", SNPs_risk)

# Save back to the same file
saveWorkbook(
  wb,
  "MS_Risk_SNPs.xlsx",
  overwrite = TRUE
)
```

## Gene Annotation by SNPNexus

We use the SNP annotation tool SNP Nexus to annotate each variant. We used overlapped gene information and related annotations.

```{r}
SNPs <- read_excel("MS_Risk_SNPs.xlsx", sheet = "SNPs_risk")

# To get it into the format for input into SNPNexus:

SNPs$BP_EXP <- as.numeric(as.character(SNPs$BP_EXP))

regions <- data.frame(
  region = "region",
  chr = SNPs$CHR_EXP,
  start = SNPs$BP_EXP,
  end = SNPs$BP_EXP
)

write.table(
  regions,
  file = "snpnexus_inputregions.txt",
  quote = FALSE,
  sep = "\t",
  row.names = FALSE,
  col.names = FALSE
)

# submit for gene annotation using UCSC annotations. Download results by "TXT per annotation" to obtain overlapped gene info. 

near_genes <- read.delim("annotation_MS_txt/near_gens.txt", stringsAsFactors = FALSE)

near_genes_collapsed <- near_genes %>%
  group_by(Chromosome, Position) %>%
  summarise(
    Overlapped.Gene = paste(unique(Overlapped.Gene), collapse = "; "),
    Type            = paste(unique(Type), collapse = "; "),
    Annotation      = paste(unique(Annotation), collapse = "; "),
    .groups = "drop"
  ) %>%
  mutate(Chromosome = sub("^chr", "", Chromosome))

SNPs_annotated <- SNPs %>%
  left_join(
    near_genes_collapsed,
    by = c("CHR_EXP" = "Chromosome",
           "BP_EXP"  = "Position")
  )

wb <- loadWorkbook("MS_Risk_SNPs.xlsx")
addWorksheet(wb, "SNPs_annotated")
writeData(wb, "SNPs_annotated", SNPs_annotated)

# Save back to the same file
saveWorkbook(
  wb,
  "MS_Risk_SNPs.xlsx",
  overwrite = TRUE
)
```

So far, 793 SNPs are missing frequency information (not in 1000G phase3 database) so they do not have risk allele information. This is 793/6150 SNPs = 13% of the SNPs. To fill in the 793 SNPs that were missing, we will use the 1000G 30X database for those that are in that database.

## Using plink with 1000G 30X for missing frequencies

From here on, the code logic is the same as previous sections. We first get a dataframe of all the SNPs that were missing (793 SNPs), separate them by ancestry to fill in frequency information, and fill in risk allele information.

The code blocks below are the same as the workflow above, just adjusted to fit the need to use the 30X coverage database.

```{r}
SNPs_missing <- read_excel("MS_Risk_SNPs.xlsx", sheet = "missing_frq")

EUR_snps <- SNPs %>% 
  filter(Ancestry == "EUR", !is.na(SNP_EXP)) %>% 
  pull(SNP_EXP) %>% 
  unique()

write.table(EUR_snps, file = "EUR_snps_of_interest_missing.txt", quote = FALSE, row.names = FALSE, col.names = FALSE)

AFR_snps <- SNPs %>% 
  filter(Ancestry == "AFR",!is.na(SNP_EXP)) %>% 
  pull(SNP_EXP) %>% 
  unique()

write.table(AFR_snps, file = "AFR_snps_of_interest_missing.txt", quote = FALSE, row.names = FALSE, col.names = FALSE)

# these files will be moved to the PLINK folder for plink --freq functions. 
```

``` unix
for chr in {1..22}; do
  plink \
    --bfile /data/CAGE_clusterdata/PLINK_1000_Genome_30X_hg38/Binary/${chr} \
    --keep EUR_keep.txt \
    --extract EUR_snps_of_interest_missing.txt \
    --freq \
    --out missing_eur_freq_chr${chr}
done

for chr in {1..22}; do
  plink \
    --bfile /data/CAGE_clusterdata/PLINK_1000_Genome_30X_hg38/Binary/${chr} \
    --keep AFR_keep.txt \
    --extract AFR_snps_of_interest_missing.txt \
    --freq \
    --out missing_afr_freq_chr${chr}
done
```

```{r}
eur_freq <- lapply(c(1:8, 10:22), function(chr) {
  read_table(
    paste0("../plink/missing_eur_freq_chr", chr, ".frq"),
    col_types = cols(
      CHR      = col_integer(),
      SNP      = col_character(),
      A1       = col_character(),
      A2       = col_character(),
      MAF      = col_double(),
      NCHROBS  = col_double()
    )
  )
}) %>%
  bind_rows()

eur_freq <- eur_freq %>%
  mutate(
    minor_allele = A1,
    major_allele = A2,
    minor_af = MAF,
    major_af = 1 - MAF
  )
eur_freq <- subset(eur_freq, select = -c(A1, A2, MAF, NCHROBS))
eur_freq_unique <- eur_freq %>%
  distinct(SNP, .keep_all = TRUE)

afr_freq <- lapply(5, function(chr) {
  read_table(
    paste0("../plink/missing_afr_freq_chr", chr, ".frq"),
    col_types = cols(
      CHR      = col_integer(),
      SNP      = col_character(),
      A1       = col_character(),
      A2       = col_character(),
      MAF      = col_double(),
      NCHROBS  = col_double()
    )
  )
}) %>%
  bind_rows()

afr_freq <- afr_freq %>%
  mutate(
    minor_allele = A1,
    major_allele = A2,
    minor_af = MAF,
    major_af = 1 - MAF
  )
afr_freq <- subset(afr_freq, select = -c(A1, A2, MAF, NCHROBS))

```

```{r}

SNPs_missing <- SNPs_missing %>% 
  left_join( 
    eur_freq %>% 
      select(SNP, major_allele, minor_allele, major_af, minor_af), 
    by = c("SNP_EXP" = "SNP") ) %>% 
  mutate( 
    snp_maj = ifelse(Ancestry == "EUR", major_allele, snp_maj), 
    snp_min = ifelse(Ancestry == "EUR", minor_allele, snp_min), 
    snp_maj_af = ifelse(Ancestry == "EUR", major_af, snp_maj_af), 
    snp_min_af = ifelse(Ancestry == "EUR", minor_af, snp_min_af) ) %>% 
  select(-major_allele, -minor_allele, -major_af, -minor_af)

SNPs_missing <- SNPs_missing %>% 
  left_join( 
    afr_freq %>% 
      select(SNP, major_allele, minor_allele, major_af, minor_af), 
    by = c("SNP_EXP" = "SNP") ) %>% 
  mutate( 
    snp_maj = ifelse(Ancestry == "AFR", major_allele, snp_maj), 
    snp_min = ifelse(Ancestry == "AFR", minor_allele, snp_min), 
    snp_maj_af = ifelse(Ancestry == "AFR", major_af, snp_maj_af), 
    snp_min_af = ifelse(Ancestry == "AFR", minor_af, snp_min_af) ) %>% 
  select(-major_allele, -minor_allele, -major_af, -minor_af)


# Take care of duplicate SNPs in dup_snps manually in Excel (we kept the duplicate with frequency that was closest to the corresponding tag SNP)
dup_snps <- SNPs_missing %>% count(SNP_EXP) %>% filter(n > 1)


SNPs_missing <- SNPs_missing %>%
  mutate(
    snp_risk = case_when(
      is.na(tag_risk) ~ "ERROR: no reported tag risk allele",
      tag_risk == tag_maj ~ snp_maj,
      tag_risk == tag_min ~ snp_min,
      TRUE ~ "ERROR: risk allele does not match"
    )
  )

wb <- loadWorkbook("MS_Risk_SNPs.xlsx")
addWorksheet(wb, "SNPS_missing_filled")
writeData(wb, "SNPS_missing_filled", SNPs_missing)

# Save back to the same file
saveWorkbook(
  wb,
  "MS_Risk_SNPs.xlsx",
  overwrite = TRUE
)
```

```{r}
# filling in the main sheet using the filled in "missing frequency" SNPs 
SNPs <- read_excel("MS_Risk_SNPs.xlsx", sheet = "SNPs_annotated")
SNPs_missing <- read_excel("MS_Risk_SNPs.xlsx", sheet = "SNPS_missing_filled")

SNPs_missing <- SNPs_missing %>%
  mutate(
    BP_EXP = as.numeric(BP_EXP),
    CHR_EXP = as.numeric(CHR_EXP),
    POS_TAG = as.numeric(POS_TAG),
    CHR_TAG = as.numeric(CHR_TAG)
  )
SNPs <- SNPs %>%
  mutate(
    BP_EXP = as.numeric(BP_EXP),
    CHR_EXP = as.numeric(CHR_EXP),
    POS_TAG = as.numeric(POS_TAG),
    CHR_TAG = as.numeric(CHR_TAG)
  )

SNPs <- SNPs %>%
  left_join(
    SNPs_missing %>%
      select(SNP_EXP, CHR_EXP, BP_EXP, SNP_TAG, CHR_TAG, POS_TAG,
             snp_maj, snp_min,
             snp_maj_af, snp_min_af,
             snp_risk),
    by = c("SNP_EXP", "CHR_EXP", "BP_EXP", "SNP_TAG"),
    suffix = c("", "_new")
  ) %>%
  mutate(
    snp_maj    = coalesce(snp_maj, snp_maj_new),
    snp_min    = coalesce(snp_min, snp_min_new),
    snp_maj_af = coalesce(snp_maj_af, snp_maj_af_new),
    snp_min_af = coalesce(snp_min_af, snp_min_af_new),
    snp_risk   = coalesce(snp_risk, snp_risk_new)
  ) %>%
  select(-ends_with("_new"))

wb <- loadWorkbook("MS_Risk_SNPs.xlsx")
addWorksheet(wb, "SNPS_filledwith_1000G30x")
writeData(wb, "SNPS_filledwith_1000G30x", SNPs)

# Save back to the same file
saveWorkbook(
  wb,
  "MS_Risk_SNPs.xlsx",
  overwrite = TRUE
)
```

-   After referencing the 1000G 30X database, 332/6150 (\~5%) SNPs are still missing frequency.
-   By the end of this workflow, we now have a list of LD-expanded MS risk variants with corresponding risk allele annotation for 95% of the variants. These risk variants are also annotated by overlapped gene and characteristics (i.e., intronic, coding, etc.)
